Implications of uniformly distributed, empirically informed priors for phylogeographical model selection: A reply to Hickerson et al.

### 2014-02-22 manuscripts/reply/reply-main.tex
Adding Intro section title to reply manuscript.
Using `\url` for all URLs in manuscript.
A few edits to reply ms.

@@

Even if we assume the gene divergence times are estimated without error,
Hickerson et al.'s \citeyear{Hickerson2013} argument only applies to one of the
three different priors used by \citet{Oaks2012}.
The narrowest prior on divergence times used by \citet{Oaks2012} ($U(0, 5
mya)$) closely mirrors the range of estimates of gene-divergence times.
Applying Hickerson et al.'s (\citeyear{Hickerson2013}) sampling-probability
argument demonstrates this prior is densely populated with samples with large
numbers of divergence parameters with values younger than the estimated gene
divergence estimates.
Thus, if insufficient prior sampling is to blame for the bias, it should be
much reduced under the narrow prior on \divt{}.
However, the magnitude of the bias is very similar across all three priors
explored by \citet{Oaks2012}.
\citet{Hickerson2013} point out a case where the narrow prior performs
slightly better (panel L of Figures S32, S37, and S38 of \citet{Oaks2012}).
However, it is important to note that these results suffered from a bug
in \msb, and there are many cases after \citet{Oaks2012} corrected the 
bug where the narrow prior performs slightly worse (see panels D--J of
Figures 3 and S12).

@@

To explore this prediction further, we repeat the analysis of the Philippines
dataset under the intermediate prior used by \citet{Oaks2012} ($\divt{} \sim
U(0, 10)$, $\meanDescendantTheta{} \sim (0.0005, 0.04)$, $\ancestralTheta{}
\sim (0.0005, 0.02)$), using a very large prior sample size of $10^8$.
When we look at the trace of the estimates of the dispersion index of
divergence times (\vmratio{}) as the prior samples accumulate
(Figure~\ref{figSamplingError}) we do not see the trend predicted by
\citet{Hickerson2013} in either the unadjusted or GLM-regression-adjusted
estimates.
While sampling error is always a reality, it does not appear to be playing a
large role in the results of \citet{Oaks2012} or presented above.

@@

This is further demonstrated by the results presented herein that show the
model-averaging approach of \citet{Hickerson2013} prefers models with narrower
\divt{} priors (Table~\ref{tabModelChoiceEmpirical} and
Figs.~\labelcref{figExclusionSimTau,figExclusionSimProb,figPowerExclusion4})
and models with fewer \divt{} parameters (Figure~\ref{figPower4}).
In all of these analyses, each of the prior models have the same number of
samples.
Thus, while sampling error will always be present in any numerical Bayesian
approximation method, insufficient sampling is an unlikely explanation for
support of prior models with less parameter space.
While analyses that sample each model proportional to their parameter space
could be explored, it is clear that the marginal likelihoods under broad
uniform priors on divergence times will be greater for models with fewer
divergence-time dimensions.

@@

The work presented herein follows the principles of Open Notebook Science.
All aspects of the work were recorded in real-time via version-control software
and are publicly available at
\href{https://github.com/joaks1/msbayes-experiments}{https://github.com/joaks1/msbayes-experiments}.
All information necessary to reproduce our results is provided there.

>>>>

\section{Introduction}

@@

Even if we assume the gene divergence times are estimated without error,
Hickerson et al.'s \citeyear{Hickerson2013} argument only applies to one of the
three different priors used by \citet{Oaks2012}.
The narrowest prior on divergence times used by \citet{Oaks2012} ($U(0, 5)$)
closely mirrors the range of estimates of gene-divergence times if we assume
the same mutation rate and one generation per year (0--5 million years ago).
Applying Hickerson et al.'s (\citeyear{Hickerson2013}) sampling-probability
argument demonstrates this prior is densely populated with samples with large
numbers of divergence parameters with values younger than the estimated gene
divergence estimates.
Thus, if insufficient prior sampling is to blame for the bias, it should be
much reduced under the narrow prior on \divt{}.
However, the magnitude of the bias is very similar across all three priors
explored by \citet{Oaks2012}.
\citet{Hickerson2013} point out a case where the narrow prior performs
slightly better (panel L of Figures S32, S37, and S38 of \citet{Oaks2012}).
However, it is important to note that these results suffered from a bug
in \msb, and there are many cases after \citet{Oaks2012} corrected the 
bug where the narrow prior performs slightly worse (see panels D--J of
Figures 3 and S12).

@@

To explore this prediction further, we repeat the analysis of the Philippines
dataset under the intermediate prior used by \citet{Oaks2012} ($\divt{} \sim
U(0, 10)$, $\meanDescendantTheta{} \sim (0.0005, 0.04)$, $\ancestralTheta{}
\sim (0.0005, 0.02)$), using a very large prior sample size of $10^8$.
When we look at the trace of the estimates of the dispersion index of
divergence times (\vmratio{}) as the prior samples accumulate
(Figure~\ref{figSamplingError}) we do not see the trend predicted by
\citet{Hickerson2013} in either the unadjusted or GLM-regression-adjusted
estimates.
While sampling error is always a reality, it does not appear to be playing a
large role in the biases revealed by the results of \citet{Oaks2012} or
presented above.

@@

This is further demonstrated by the results presented herein that show the
model-averaging approach of \citet{Hickerson2013} prefers models with narrower
\divt{} priors (Table~\ref{tabModelChoiceEmpirical} and
Figs.~\labelcref{figExclusionSimTau,figExclusionSimProb,figPowerExclusion4})
and models with fewer \divt{} parameters (Figure~\ref{figPower4}).
In all of these analyses, each of the prior models have the same number of
samples.
Thus, while sampling error will always be present in any numerical Bayesian
approximation method, insufficient sampling is an unlikely explanation for the
support of prior models with less parameter space.
While analyses that sample each model proportional to their parameter space
could be explored, it is clear that the marginal likelihoods under broad
uniform priors on divergence times will be greater for models with fewer
divergence-time dimensions.

@@

The work presented herein follows the principles of Open Notebook Science.
All aspects of the work were recorded in real-time via version-control software
and are publicly available at
\href{https://github.com/joaks1/msbayes-experiments}{\url{https://github.com/joaks1/msbayes-experiments}}.
All information necessary to reproduce our results is provided there.

### 2014-02-22 manuscripts/reply/reply.tex
A few edits to reply ms.

\begin{abstract}
    Biogeographers frequently seek to explain population and species
    differentiation on geographical phenomena.
    Establishing that a set of population splitting events occurred
    at the same time can be a potentially persuasive argument that a set of taxa
    were affected by the same geographic events.
    Unfortunately, estimating divergence times precisely when one lacks
    precise information about the rate of molecular evolution is a notoriously
    difficult problem in evolutionary biology.
    \citet{Huang2011} introduced an approximate Bayesian approach which can be
    used to estimate the probabilities of models in which multiple sets of taxa
    diverge at the same time. 
    This method (implemented in the software \msb) does not require a common
    set of character data that links all of the taxa under investigation, but
    does require the relative mutation rates among the taxa and loci to be
    assumed \emph{a priori}.
    Recently, \citet{Oaks2012} used this model-choice framework to 
    study 22 pairs of vertebrate lineages which are distributed across
    the Philippines; they also studied the behavior of the \msb approach
    using computer simulations.
    \citet{Oaks2012} found that the model was very sensitive to the prior and
    that the inference method had low power to detect variation in divergences
    times.
    These results were not surprising, in light of a rich statistical literature
    showing that the marginal likelihood of a model is sensitive to the
    use of vague priors.
    Because this sensitivity to prior assumptions affects the crucial insights 
    that a researcher who employs \msb seeks to gain, \citet{Oaks2012} recommended
    that users of the approach should carefully assess the robustness of their 
    conclusions to different priors.
    According to \citet{Hickerson2013}, the lack of robustness in \msb analyses
    was due to excessively broad priors on divergence times, leading to 
    inadequate numbers of simulation replicates.
    They proposed a new model-averaging approach that uses narrow, empirically
    informed uniform priors.
    Here we demonstrate that the approach of \citet{Hickerson2013} is
    dangerous in the sense that the empirically-derived priors often
    exclude from consideration the true values of the models' parameters.
    On a more fundamental level, we question the value of adopting an empirical
    Bayesian stance for this model-choice problem, because it can mislead model
    posterior probabilities, which are inherently measures of belief in the
    models after prior knowledge is updated by the data.
    The robust Bayesian approach of conducting analyses under a variety
    of priors can reveal prior sensitivity and communicate which assumptions
    underlie model inference.
    Furthermore, simulations provide insight into the temporal resolution of
    the method, which in turn helps guide interpretation of results.

>>>>

\begin{abstract}
    Biogeographers frequently seek to explain population and species
    differentiation on geographical phenomena.
    Establishing that a set of population splitting events occurred
    at the same time can be a potentially persuasive argument that a set of taxa
    were affected by the same geographic events.
    Unfortunately, estimating divergence times precisely when one lacks
    precise information about the rate of molecular evolution is a notoriously
    difficult problem in evolutionary biology.
    \citet{Huang2011} introduced an approximate Bayesian approach (implemented
    in the software \msb) which can be used to estimate the probabilities of
    models in which multiple sets of taxa diverge at the same time. 
    % This method  does not require a common set of character data that links all
    % of the taxa under investigation, but does require the relative mutation
    % rates among the taxa and loci to be assumed \emph{a priori}.
    Recently, \citet{Oaks2012} used this model-choice framework to 
    study 22 pairs of vertebrate lineages which are distributed across
    the Philippines; they also studied the behavior of the \msb approach
    using computer simulations.
    \citet{Oaks2012} found that the model was very sensitive to the prior and
    that the inference method had low power to detect variation in divergences
    times.
    These results were not surprising, in light of a rich statistical literature
    showing that the marginal likelihood of a model is sensitive to the
    use of vague priors.
    Because this sensitivity to prior assumptions affects the crucial insights 
    that a researcher who employs \msb seeks to gain, \citet{Oaks2012} recommended
    that users of the approach should carefully assess the robustness of their 
    conclusions to different priors.
    According to \citet{Hickerson2013}, the lack of robustness in \msb analyses
    was due to excessively broad priors on divergence times, leading to 
    inadequate numbers of simulation replicates.
    They proposed a new model-averaging approach that uses narrow, empirically
    informed uniform priors.
    Here we demonstrate that the approach of \citet{Hickerson2013} is
    dangerous in the sense that the empirically-derived priors often
    exclude from consideration the true values of the models' parameters.
    On a more fundamental level, we question the value of adopting an empirical
    Bayesian stance for this model-choice problem, because it can mislead model
    posterior probabilities, which are inherently measures of belief in the
    models after prior knowledge is updated by the data.
    The robust Bayesian approach of conducting analyses under a variety
    of priors can reveal prior sensitivity and communicate which assumptions
    underlie model inference.
    Furthermore, simulations provide insight into the temporal resolution of
    the method, which in turn helps guide interpretation of results.
